---
title: "R Mailing List Data"
format: gfm
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(nanoparquet)
library(dplyr, warn.conflicts = FALSE)
library(scales)
library(ggplot2)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot"
    )
)

read_messages <- function(col_select = NULL) {
  files <- list.files("data/messages", pattern = "\\.parquet$", full.names = TRUE)
  do.call(rbind, lapply(files, read_parquet, col_select = col_select))
}
```

A collection of [R mailing list](https://www.r-project.org/mail.html) archives in [Parquet](https://parquet.apache.org/) format, ready for analysis in R, Python, or any language with Parquet support.

Data is sourced from the [R Mailing Lists archive project](https://github.com/r-mailing-lists) and updated automatically. You can also browse the archives at [r-mailing-lists.thecoatlessprofessor.com](https://r-mailing-lists.thecoatlessprofessor.com).

## Quick start

Copy the helper functions below to download and read the data directly from GitHub — no cloning required. Files are cached locally so repeated calls are fast.

### R

```r
# -- Helper functions (copy once) ---------------------------------------------
# Dependencies: jsonlite, nanoparquet

.rml_base_url <- "https://raw.githubusercontent.com/r-mailing-lists/data/main/data"
.rml_api_url  <- "https://api.github.com/repos/r-mailing-lists/data/contents/data/messages"
.rml_cache    <- file.path(tempdir(), "rml_cache")

.rml_ensure_cache <- function() {
  if (!dir.exists(.rml_cache)) dir.create(.rml_cache, recursive = TRUE)
  .rml_cache
}

.rml_download <- function(url, destfile) {
  .rml_ensure_cache()
  if (!file.exists(destfile)) {
    message("Downloading ", basename(destfile), "...")
    download.file(url, destfile, mode = "wb", quiet = TRUE)
  }
  destfile
}

.rml_file_index <- function() {
  cache_file <- file.path(.rml_ensure_cache(), "_index.json")
  if (!file.exists(cache_file)) {
    message("Fetching file index...")
    json <- jsonlite::fromJSON(.rml_api_url)
    jsonlite::write_json(json, cache_file)
  }
  jsonlite::fromJSON(cache_file)
}

rml_available <- function() {
  index <- .rml_file_index()
  files <- sub("\\.parquet$", "", index$name)
  sort(unique(sub("-\\d{4}-\\d{4}$", "", files)))
}

rml_read <- function(list_name, col_select = NULL) {
  index <- .rml_file_index()
  pattern <- paste0("^", list_name, "(-\\d{4}-\\d{4})?\\.parquet$")
  matches <- index$name[grepl(pattern, index$name)]
  if (length(matches) == 0) {
    stop("List '", list_name, "' not found. Run rml_available() to see options.",
         call. = FALSE)
  }
  frames <- lapply(matches, function(f) {
    dest <- file.path(.rml_ensure_cache(), f)
    .rml_download(paste0(.rml_base_url, "/messages/", f), dest)
    nanoparquet::read_parquet(dest, col_select = col_select)
  })
  if (length(frames) == 1) frames[[1]] else do.call(rbind, frames)
}

rml_read_threads <- function(col_select = NULL) {
  dest <- file.path(.rml_ensure_cache(), "threads.parquet")
  .rml_download(paste0(.rml_base_url, "/threads.parquet"), dest)
  nanoparquet::read_parquet(dest, col_select = col_select)
}

rml_read_contributors <- function(col_select = NULL) {
  dest <- file.path(.rml_ensure_cache(), "contributors.parquet")
  .rml_download(paste0(.rml_base_url, "/contributors.parquet"), dest)
  nanoparquet::read_parquet(dest, col_select = col_select)
}
```

```r
# See available lists
rml_available()

# Top 10 r-devel posters in the last year
msgs <- rml_read("r-devel", col_select = c("from_name", "date"))
recent <- msgs[msgs$date >= as.POSIXct(Sys.Date() - 365), ]
head(sort(table(recent$from_name), decreasing = TRUE), 10)

# Message counts per list (uses thread summaries — single small download)
threads <- rml_read_threads(col_select = c("list", "message_count"))
agg <- aggregate(message_count ~ list, data = threads, FUN = sum)
agg[order(-agg$message_count), ]
```

### Python

```python
# -- Helper functions (copy once) ---------------------------------------------
# Dependencies: polars

import json
import re
import tempfile
from pathlib import Path
from urllib.request import urlopen, urlretrieve

import polars as pl

_RAW_BASE = "https://raw.githubusercontent.com/r-mailing-lists/data/main/data"
_API_URL = "https://api.github.com/repos/r-mailing-lists/data/contents/data/messages"
_CACHE_DIR = Path(tempfile.gettempdir()) / "rml_cache"
_DECADE_RE = re.compile(r"^(.+)-(\d{4})-(\d{4})\.parquet$")


def _cached_download(url: str, filename: str) -> Path:
    _CACHE_DIR.mkdir(parents=True, exist_ok=True)
    dest = _CACHE_DIR / filename
    if not dest.exists():
        urlretrieve(url, dest)
    return dest


def _list_message_files() -> list[dict]:
    with urlopen(_API_URL) as resp:
        return json.loads(resp.read())


def rml_available() -> list[str]:
    files = _list_message_files()
    names: set[str] = set()
    for f in files:
        m = _DECADE_RE.match(f["name"])
        names.add(m.group(1) if m else f["name"].removesuffix(".parquet"))
    return sorted(names)


def rml_read(list_name: str) -> pl.DataFrame:
    files = _list_message_files()
    matched = [
        f for f in files
        if f["name"] == f"{list_name}.parquet"
        or (_DECADE_RE.match(f["name"]) and
            _DECADE_RE.match(f["name"]).group(1) == list_name)
    ]
    if not matched:
        raise ValueError(f"List '{list_name}' not found. See rml_available()")
    dfs = [
        pl.read_parquet(_cached_download(f["download_url"], f["name"]))
        for f in matched
    ]
    return pl.concat(dfs) if len(dfs) > 1 else dfs[0]


def rml_read_threads() -> pl.DataFrame:
    return pl.read_parquet(
        _cached_download(f"{_RAW_BASE}/threads.parquet", "threads.parquet")
    )


def rml_read_contributors() -> pl.DataFrame:
    return pl.read_parquet(
        _cached_download(f"{_RAW_BASE}/contributors.parquet", "contributors.parquet")
    )
```

```python
# See available lists
rml_available()

# Top 10 r-devel posters in 2025
df = rml_read("r-devel")
(df.filter(pl.col("date").dt.year() == 2025)
   .group_by("from_name")
   .len()
   .sort("len", descending=True)
   .head(10))

# Message counts per list (uses thread summaries — single small download)
(rml_read_threads()
   .group_by("list")
   .agg(pl.col("message_count").sum())
   .sort("message_count", descending=True))
```

### Working with a local clone

If you prefer to work with local files, clone the repo and read the parquet files directly:

```r
# R
library(nanoparquet)
r_devel <- read_parquet("data/messages/r-devel.parquet")

# Read all lists into one data frame
files <- list.files("data/messages", pattern = "\\.parquet$", full.names = TRUE)
all_msgs <- do.call(rbind, lapply(files, read_parquet))
```

```python
# Python
import polars as pl
r_devel = pl.read_parquet("data/messages/r-devel.parquet")
all_msgs = pl.read_parquet("data/messages/*.parquet")
```

## Data overview

```{r}
#| label: summary-table
#| results: asis
msgs <- read_messages(col_select = c("list", "from_name", "date", "month"))

summary_df <- msgs |>
  group_by(list) |>
  summarise(
    messages = n(),
    authors = n_distinct(from_name),
    first_message = as.Date(min(date, na.rm = TRUE)),
    last_message = as.Date(max(date, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(desc(messages))

cat(paste0(
  "**", format(sum(summary_df$messages), big.mark = ","),
  "** messages across **", nrow(summary_df), "** mailing lists\n"
))

summary_df |>
  mutate(
    messages = format(messages, big.mark = ","),
    authors = format(authors, big.mark = ","),
    first_message = format(first_message, "%b %Y"),
    last_message = format(last_message, "%b %Y")
  ) |>
  knitr::kable(col.names = c("List", "Messages", "Authors", "First Message", "Last Message"))
```

```{r}
#| label: fig-timeline
#| fig-alt: "Monthly message volume across R mailing lists over time"
#| fig-width: 10
#| fig-height: 5
monthly <- msgs |>
  filter(date >= as.POSIXct("1997-01-01", tz = "UTC")) |>
  count(list, month) |>
  mutate(date = as.Date(paste0(month, "-01")))

top_lists <- monthly |>
  group_by(list) |>
  summarise(total = sum(n)) |>
  slice_max(total, n = 5) |>
  pull(list)

monthly |>
  filter(list %in% top_lists) |>
  ggplot(aes(date, n, color = list)) +
  geom_line(alpha = 0.7, linewidth = 0.5) +
  geom_smooth(se = FALSE, linewidth = 1, span = 0.15) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Monthly message volume (top 5 lists)",
    x = NULL, y = "Messages per month", color = "List"
  )
```

## Example: Reply network on r-devel

The `in_reply_to` field links each message to its parent, making it straightforward to build a "who replies to whom" network.

```{r}
#| label: fig-reply-network
#| fig-alt: "Network graph showing reply relationships between top r-devel contributors"
#| fig-width: 10
#| fig-height: 10
library(igraph)
library(ggraph)

r_devel <- read_parquet(
  "data/messages/r-devel.parquet",
  col_select = c("message_id", "from_name", "in_reply_to")
)

# Build edges: replier -> original author
author_lookup <- r_devel |> select(message_id, from_name)

edges <- r_devel |>
  filter(!is.na(in_reply_to)) |>
  inner_join(author_lookup, by = c("in_reply_to" = "message_id"), suffix = c("_from", "_to")) |>
  filter(from_name_from != from_name_to) |>
  count(from = from_name_from, to = from_name_to, name = "replies")

# Keep only the most active participants
top_authors <- r_devel |>
  count(from_name, sort = TRUE) |>
  head(30) |>
  pull(from_name)

edges_top <- edges |>
  filter(from %in% top_authors, to %in% top_authors, replies >= 5)

g <- graph_from_data_frame(edges_top, directed = TRUE)

# Size nodes by total messages
msg_counts <- r_devel |>
  filter(from_name %in% V(g)$name) |>
  count(from_name)
V(g)$messages <- msg_counts$n[match(V(g)$name, msg_counts$from_name)]

ggraph(g, layout = "fr") +
  geom_edge_link(
    aes(width = replies, alpha = replies),
    arrow = arrow(length = unit(2, "mm"), type = "closed"),
    end_cap = circle(4, "mm")
  ) +
  geom_node_point(aes(size = messages), color = "#3B6EA8") +
  geom_node_text(aes(label = name), repel = TRUE, size = 3, max.overlaps = 20) +
  scale_edge_width(range = c(0.3, 2.5), guide = "none") +
  scale_edge_alpha(range = c(0.15, 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 12), labels = label_comma(), name = "Messages") +
  labs(title = "Reply network among top r-devel contributors") +
  theme_void(base_size = 13) +
  theme(plot.title.position = "plot", legend.position = "bottom")
```

## Data dictionary

### `data/messages/<list>.parquet`

One Parquet file per mailing list. All files share the same schema.

| Column | Type | Description |
|--------|------|-------------|
| `list` | string | Mailing list name (e.g., `r-devel`) |
| `id` | string | Unique message ID (`msg-<hash>`) |
| `message_id` | string | Original RFC 2822 Message-ID header |
| `from_name` | string | Author display name |
| `from_email_hash` | string | SHA-256 hash of author email (privacy-preserving) |
| `date` | timestamp | Message date (UTC) |
| `subject` | string | Subject line with `Re:`/`Fwd:` prefixes stripped |
| `in_reply_to` | string | ID of the parent message (null for thread starters) |
| `body` | string | Full message body text |
| `body_snippet` | string | First 200 characters of the body |
| `thread_id` | string | Thread grouping ID (`thread-<hash>`) |
| `thread_depth` | integer | Depth in thread tree (0 = root message) |
| `month` | string | `YYYY-MM` for temporal bucketing |

### `data/threads.parquet`

Thread-level summaries for all lists.

| Column | Type | Description |
|--------|------|-------------|
| `list` | string | Mailing list name |
| `id` | string | Thread ID (`thread-<hash>`) |
| `subject` | string | Thread subject |
| `message_count` | integer | Number of messages in thread |
| `started` | timestamp | Date of first message |
| `last_reply` | timestamp | Date of most recent reply |
| `root_message_id` | string | ID of the thread-starting message |

### `data/contributors.parquet`

Aggregated contributor statistics across all lists.

| Column | Type | Description |
|--------|------|-------------|
| `name` | string | Author display name |
| `message_count` | integer | Total messages across all lists |
| `list_count` | integer | Number of distinct lists posted to |
| `lists` | string | Comma-separated list names |

## Privacy

Email addresses are not included in this dataset. Author identity is represented by display name and a SHA-256 hash of the email address, which allows grouping messages by author without exposing contact information. The original emails are publicly archived on the source mailing list servers.

## License

The mailing list content is publicly archived by the [R Project](https://www.r-project.org/mail.html) via [ETH Zurich](https://stat.ethz.ch/pipermail/) and [R-Forge](https://lists.r-forge.r-project.org/pipermail/). This dataset reformats that public content for easier analysis. The tooling in this repository is licensed under the [MIT License](LICENSE).
