---
title: "R Mailing List Data"
format: gfm
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(nanoparquet)
library(dplyr, warn.conflicts = FALSE)

read_messages <- function(col_select = NULL) {
  files <- list.files("data/messages", pattern = "\\.parquet$", full.names = TRUE)
  do.call(rbind, lapply(files, read_parquet, col_select = col_select))
}

# Source helpers and point at local data (avoids downloading during render)
source("scripts/rml.R")
.rml_data_dir <- "data"
```

A collection of [R mailing list](https://www.r-project.org/mail.html) archives in [Parquet](https://parquet.apache.org/) format, ready for analysis in R, Python, or any language with Parquet support.

Data is sourced from the [R Mailing Lists archive project](https://github.com/r-mailing-lists) and updated automatically. You can also browse the archives at [r-mailing-lists.thecoatlessprofessor.com](https://r-mailing-lists.thecoatlessprofessor.com).

## Quick start

### R

Requires [`jsonlite`](https://cran.r-project.org/package=jsonlite) and [`nanoparquet`](https://cran.r-project.org/package=nanoparquet). Source the helper script directly from GitHub:

```r
source("https://raw.githubusercontent.com/r-mailing-lists/data/main/scripts/rml.R")
```

```r
# See available lists
rml_available()

# Read only metadata columns (skips body text — much faster)
r_devel <- rml_read("r-devel",
  col_select = c("from_name", "date", "subject", "thread_id", "month"))

# Top 10 posters in the last year
recent <- r_devel[r_devel$date >= as.POSIXct(Sys.Date() - 365), ]
head(sort(table(recent$from_name), decreasing = TRUE), 10)

# Message counts per list (thread summaries — single small download)
threads <- rml_read_threads(col_select = c("list", "message_count"))
aggregate(message_count ~ list, data = threads, FUN = sum)

# Top contributors across all lists
contribs <- rml_read_contributors()
head(contribs[order(-contribs$message_count), ], 10)
```

```{r}
#| label: r-load-data
#| message: false
r_devel <- rml_read("r-devel",
  col_select = c("from_name", "date", "subject", "thread_id", "month"))
threads <- rml_read_threads(col_select = c("list", "message_count"))
contribs <- rml_read_contributors()
```

```{r}
#| label: r-summary-stats
#| results: asis
cat(paste0(
  "The `r-devel` list has **",
  format(nrow(r_devel), big.mark = ","),
  "** messages. Top 10 contributors across all lists:\n\n"
))

contribs |>
  dplyr::arrange(desc(message_count)) |>
  head(10) |>
  dplyr::select(name, message_count, list_count) |>
  dplyr::mutate(message_count = format(message_count, big.mark = ",")) |>
  knitr::kable(col.names = c("Name", "Messages", "Lists"))
```

### Python

Requires [`polars`](https://pola.rs/). Download the helper script:

```bash
curl -O https://raw.githubusercontent.com/r-mailing-lists/data/main/scripts/rml.py
```

```python
from rml import rml_available, rml_read, rml_read_threads, rml_read_contributors
import polars as pl

# See available lists
rml_available()

# Read a single list (downloads and caches parquet files automatically)
r_devel = rml_read("r-devel")

# Top 10 posters in 2025
(r_devel
   .filter(pl.col("date").dt.year() == 2025)
   .group_by("from_name")
   .len()
   .sort("len", descending=True)
   .head(10))

# Message counts per list (thread summaries — single small download)
(rml_read_threads()
   .group_by("list")
   .agg(pl.col("message_count").sum())
   .sort("message_count", descending=True))

# Top contributors across all lists
(rml_read_contributors()
   .sort("message_count", descending=True)
   .head(20))
```

### Working with a local clone

If you prefer working with local files, clone the repo and read parquet files directly:

```r
# R
library(nanoparquet)
r_devel <- read_parquet("data/messages/r-devel.parquet")
```

```python
# Python
import polars as pl
r_devel = pl.read_parquet("data/messages/r-devel.parquet")
all_msgs = pl.read_parquet("data/messages/*.parquet")
```

## Data overview

```{r}
#| label: summary-table
#| results: asis
msgs <- read_messages(col_select = c("list", "from_name", "date", "month"))

summary_df <- msgs |>
  group_by(list) |>
  summarise(
    messages = n(),
    authors = n_distinct(from_name),
    first_message = as.Date(min(date, na.rm = TRUE)),
    last_message = as.Date(max(date, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(desc(messages))

cat(paste0(
  "**", format(sum(summary_df$messages), big.mark = ","),
  "** messages across **", nrow(summary_df), "** mailing lists\n"
))

summary_df |>
  mutate(
    messages = format(messages, big.mark = ","),
    authors = format(authors, big.mark = ","),
    first_message = format(first_message, "%b %Y"),
    last_message = format(last_message, "%b %Y")
  ) |>
  knitr::kable(col.names = c("List", "Messages", "Authors", "First Message", "Last Message"))
```

For more in-depth analysis examples (message volume trends, reply networks, top contributors), see the [demo analysis](analysis/demo-analysis.md).

## Data dictionary

### `data/messages/<list>.parquet`

One Parquet file per mailing list. All files share the same schema.

| Column | Type | Description |
|--------|------|-------------|
| `list` | string | Mailing list name (e.g., `r-devel`) |
| `id` | string | Unique message ID (`msg-<hash>`) |
| `message_id` | string | Original RFC 2822 Message-ID header |
| `from_name` | string | Author display name |
| `from_email_hash` | string | SHA-256 hash of author email (privacy-preserving) |
| `date` | timestamp | Message date (UTC) |
| `subject` | string | Subject line with `Re:`/`Fwd:` prefixes stripped |
| `in_reply_to` | string | ID of the parent message (null for thread starters) |
| `body` | string | Full message body text |
| `body_snippet` | string | First 200 characters of the body |
| `thread_id` | string | Thread grouping ID (`thread-<hash>`) |
| `thread_depth` | integer | Depth in thread tree (0 = root message) |
| `month` | string | `YYYY-MM` for temporal bucketing |

### `data/threads.parquet`

Thread-level summaries for all lists.

| Column | Type | Description |
|--------|------|-------------|
| `list` | string | Mailing list name |
| `id` | string | Thread ID (`thread-<hash>`) |
| `subject` | string | Thread subject |
| `message_count` | integer | Number of messages in thread |
| `started` | timestamp | Date of first message |
| `last_reply` | timestamp | Date of most recent reply |
| `root_message_id` | string | ID of the thread-starting message |

### `data/contributors.parquet`

Aggregated contributor statistics across all lists.

| Column | Type | Description |
|--------|------|-------------|
| `name` | string | Author display name |
| `message_count` | integer | Total messages across all lists |
| `list_count` | integer | Number of distinct lists posted to |
| `lists` | string | Comma-separated list slugs |
| `list_counts` | string | Per-list message counts (e.g. `r-devel:150,r-help:42`) |
| `first_message` | string | ISO 8601 date of earliest message |
| `last_message` | string | ISO 8601 date of most recent message |

## Privacy

Email addresses are not included in this dataset. Author identity is represented by display name and a SHA-256 hash of the email address, which allows grouping messages by author without exposing contact information. The original emails are publicly archived on the source mailing list servers.

## License

The mailing list content is publicly archived by the [R Project](https://www.r-project.org/mail.html) via [ETH Zurich](https://stat.ethz.ch/pipermail/) and [R-Forge](https://lists.r-forge.r-project.org/pipermail/). This dataset reformats that public content for easier analysis. The tooling in this repository is licensed under the [MIT License](LICENSE).
