---
title: "R Mailing List Data: Demo Analysis"
format: gfm
execute:
  echo: true
  warning: false
---

This notebook demonstrates how to work with the R mailing list Parquet data.
See the [main README](../README.md) for setup instructions and data dictionary.

## Setup

```{r}
#| label: setup
library(nanoparquet)
library(dplyr, warn.conflicts = FALSE)
library(scales)
library(ggplot2)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot"
    )
)

read_messages <- function(col_select = NULL) {
  files <- list.files("../data/messages", pattern = "\\.parquet$", full.names = TRUE)
  do.call(rbind, lapply(files, read_parquet, col_select = col_select))
}
```

## Message volume over time

```{r}
#| label: fig-timeline
#| fig-alt: "Monthly message volume across R mailing lists over time"
#| fig-width: 10
#| fig-height: 5
msgs <- read_messages(col_select = c("list", "from_name", "date", "month"))

monthly <- msgs |>
  filter(date >= as.POSIXct("1997-01-01", tz = "UTC")) |>
  count(list, month) |>
  mutate(date = as.Date(paste0(month, "-01")))

top_lists <- monthly |>
  group_by(list) |>
  summarise(total = sum(n)) |>
  slice_max(total, n = 5) |>
  pull(list)

monthly |>
  filter(list %in% top_lists) |>
  ggplot(aes(date, n, color = list)) +
  geom_line(alpha = 0.7, linewidth = 0.5) +
  geom_smooth(se = FALSE, linewidth = 1, span = 0.15) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Monthly message volume (top 5 lists)",
    x = NULL, y = "Messages per month", color = "List"
  )
```

## Top posters by list

```{r}
#| label: top-posters
r_devel <- read_parquet(
  "../data/messages/r-devel.parquet",
  col_select = c("from_name", "date", "subject")
)

recent <- r_devel[r_devel$date >= as.POSIXct(Sys.Date() - 365), ]
head(sort(table(recent$from_name), decreasing = TRUE), 10)
```

## Reply network on r-devel

The `in_reply_to` field links each message to its parent, making it
straightforward to build a "who replies to whom" network.

```{r}
#| label: fig-reply-network
#| fig-alt: "Network graph showing reply relationships between top r-devel contributors"
#| fig-width: 10
#| fig-height: 10
library(igraph)
library(ggraph)

r_devel <- read_parquet(
  "../data/messages/r-devel.parquet",
  col_select = c("message_id", "from_name", "in_reply_to")
)

# Build edges: replier -> original author
author_lookup <- r_devel |> select(message_id, from_name)

edges <- r_devel |>
  filter(!is.na(in_reply_to)) |>
  inner_join(author_lookup, by = c("in_reply_to" = "message_id"), suffix = c("_from", "_to")) |>
  filter(from_name_from != from_name_to) |>
  count(from = from_name_from, to = from_name_to, name = "replies")

# Keep only the most active participants
top_authors <- r_devel |>
  count(from_name, sort = TRUE) |>
  head(30) |>
  pull(from_name)

edges_top <- edges |>
  filter(from %in% top_authors, to %in% top_authors, replies >= 5)

g <- graph_from_data_frame(edges_top, directed = TRUE)

# Size nodes by total messages
msg_counts <- r_devel |>
  filter(from_name %in% V(g)$name) |>
  count(from_name)
V(g)$messages <- msg_counts$n[match(V(g)$name, msg_counts$from_name)]

ggraph(g, layout = "fr") +
  geom_edge_link(
    aes(width = replies, alpha = replies),
    arrow = arrow(length = unit(2, "mm"), type = "closed"),
    end_cap = circle(4, "mm")
  ) +
  geom_node_point(aes(size = messages), color = "#3B6EA8") +
  geom_node_text(aes(label = name), repel = TRUE, size = 3, max.overlaps = 20) +
  scale_edge_width(range = c(0.3, 2.5), guide = "none") +
  scale_edge_alpha(range = c(0.15, 0.6), guide = "none") +
  scale_size_continuous(range = c(2, 12), labels = label_comma(), name = "Messages") +
  labs(title = "Reply network among top r-devel contributors") +
  theme_void(base_size = 13) +
  theme(plot.title.position = "plot", legend.position = "bottom")
```

## Contributors across lists

```{r}
#| label: top-contributors
contribs <- read_parquet("../data/contributors.parquet")
contribs |>
  arrange(desc(message_count)) |>
  head(20) |>
  select(name, message_count, list_count) |>
  knitr::kable()
```
